---
title: "HW 2 Student"
author: "Andy Ackerman"
date: "10/17/2023"
output: 
  html_document:
    number_sections: true
---

This homework is meant to illustrate the methods of classification algorithms as well as their potential pitfalls.  In class, we demonstrated K-Nearest-Neighbors using the `iris` dataset.  Today I will give you a different subset of this same data, and you will train a KNN classifier.  

```{r, echo = FALSE}
set.seed(123)
library(class)

df <- data(iris) 

normal <-function(x) {
  (x -min(x))/(max(x)-min(x))   
}

iris_norm <- as.data.frame(lapply(iris[,c(1,2,3,4)], normal))

subset <- c(1:45, 58, 60:70, 82, 94, 110:150)
iris_train <- iris_norm[subset,] 
iris_test <- iris_norm[-subset,] 

iris_target_category <- iris[subset,5]
iris_test_category <- iris[-subset,5]


```

#
Above, I have given you a training-testing partition.  Train the KNN with $K = 5$ on the training data and use this to classify the 50 test observations.  Once you have classified the test observations, create a contingency table -- like we did in class -- to evaluate which observations your algorithm is misclassifying.   

```{r}
set.seed(123)

predicted_categories <- knn(train = iris_train, test = iris_test, cl = iris_target_category, k = 5)

contingency_table <- table(Predicted = predicted_categories, Actual = iris_test_category)

print(contingency_table)
accuracy <- sum(diag(contingency_table)) / sum(contingency_table)
print(paste("Accuracy: ", accuracy))

```

#

Discuss your results.  If you have done this correctly, you should have a classification error rate that is roughly 20% higher than what we observed in class.  Why is this the case? In particular run a summary of the `iris_test_category` as well as `iris_target_category` and discuss how this plays a role in your answer.  

*STUDENT INPUT* 
In our course, we allocated the training and testing datasets randomly. However, in the assignments, the division between training and testing sets is done manually. This method of comparison against random assignment can introduce bias and an uneven data distribution. If the training dataset doesn't accurately reflect the characteristics of the testing set, particularly with regard to the distribution of classes, it can compromise the effectiveness of the classifier. For instance, if a class is disproportionately represented in the training dataset or insufficiently represented in the testing dataset, it could lead to an increased rate of errors.
#

Build a github repository to store your homework assignments.  Share the link in this file.  

*STUDENT INPUT*
https://github.com/YueweiWang2/STOR-390-HW
