---
title: "HW 4"
author: "Student Name"
date: "12/29/2023"
output: 
  html_document:
    number_sections: true
---

This homework is designed to give you practice fitting a logistic regression and working with statistical/philosophical measures of fairness.  We will work with the `titanic` dataset which we have previously seen in class in connection to decision trees.  

Below I will preprocess the data precisely as we did in class.  You can simply refer to `data_train` as your training data and `data_test` as your testing data.  




```{r}

#this is all of the preprocessing done for the decision trees lecture.  

path <- 'https://raw.githubusercontent.com/guru99-edu/R-Programming/master/titanic_data.csv'
titanic <-read.csv(path)
head(titanic)

library(dplyr)

#replace ? with NA
replace_question_mark <- function(x) {
  if (is.character(x)) {
    x <- na_if(x, "?")
  }
  return(x)
}

titanic <- titanic %>%
  mutate_all(replace_question_mark)

set.seed(678)
shuffle_index <- sample(1:nrow(titanic))
head(shuffle_index)

titanic <- titanic[shuffle_index, ]
head(titanic)

library(dplyr)
# Drop variables
clean_titanic <- titanic %>%
select(-c(home.dest, cabin, name, x, ticket)) %>% 
#Convert to factor level
    mutate(pclass = factor(pclass, levels = c(1, 2, 3), labels = c('Upper', 'Middle', 'Lower')),
    survived = factor(survived, levels = c(0, 1), labels = c('No', 'Yes'))) %>%
na.omit()
#previously were characters
clean_titanic$age <- as.numeric(clean_titanic$age)
clean_titanic$fare <- as.numeric(clean_titanic$fare)
glimpse(clean_titanic)

create_train_test <- function(data, size = 0.8, train = TRUE) {
    n_row = nrow(data)
    total_row = size * n_row
    train_sample <- 1: total_row
    if (train == TRUE) {
        return (data[train_sample, ])
    } else {
        return (data[-train_sample, ])
    }
}
data_train <- create_train_test(clean_titanic, 0.8, train = TRUE)
data_test <- create_train_test(clean_titanic, 0.8, train = FALSE)

```

#
Create a table reporting the proportion of people in the training set surviving the Titanic.  Do the same for the testing set.  Comment on whether the current training-testing partition looks suitable.  

```{r}
survival_rate_train <- table(data_train$survived) / nrow(data_train)
survival_rate_test <- table(data_test$survived) / nrow(data_test)
survival_summary <- rbind(Training = survival_rate_train, Testing = survival_rate_test)
print(survival_summary)

```

*student input*
The training-testing partition with survival rates of 39.8% in training and 44.5% in testing is suitable for predictive modeling. The slight difference in survival rates is acceptable, ensuring the model is trained and tested on comparable data, supporting effective model evaluation and generalization on the Titanic dataset.

#
Use the `glm` command to build a logistic regression on the training partition.  `survived` should be your response variable and `pclass`, `sex`, `age`, `sibsp`, and `parch` should be your response variables.  

```{r}
library(dplyr)
model1 <- glm(survived ~ pclass + sex + age + sibsp + parch, data = data_train, family = binomial)
summary(model1)

```

We would now like to test whether this classifier is *fair* across the sex subgroups.  It was reported that women and children were prioritized on the life-boats and as a result survived the incident at a much higher rate.  Let us see if our model is able to capture this fact.  

#

Subset your test data into a male group and a female group.  Then, use the `predict` function on the male testing group to come up with predicted probabilities of surviving the Titanic for each male in the testing set.  Do the same for the female testing group.  

```{r}

data_test_male <- subset(data_test, sex == 'male')
data_test_female <- subset(data_test, sex == 'female')

predicted_probs_male <- predict(model1, newdata = data_test_male, type = "response")

predicted_probs_female <- predict(model1, newdata = data_test_female, type = "response")

mean_predicted_prob_male <- mean(predicted_probs_male)
mean_predicted_prob_female <- mean(predicted_probs_female)

cat("Male:", mean_predicted_prob_male, "\n")
cat("Female:", mean_predicted_prob_female, "\n")


```

# 

Now recall that for this logistic *regression* to be a true classifier, we need to pair it with a decision boundary.  Use an `if-else` statement to translate any predicted probability in the male group greater than $0.5$ into `Yes` (as in Yes this individual is predicted to have survived).  Likewise an predicted probability less than $0.5$ should be translated into a `No`.  

Do this for the female testing group as well, and then create a confusion matrix for each of the male and female test set predictions.  You can use the `confusionMatrix` command as seen in class to expidite this process as well as provide you necessary metrics for the following questions.  

```{r}
library(ggplot2)
library(caret)

predicted_class_male <- ifelse(predicted_probs_male > 0.5, 'Yes', 'No')

predicted_class_female <- ifelse(predicted_probs_female > 0.5, 'Yes', 'No')
conf_matrix_male <- confusionMatrix(factor(predicted_class_male), factor(data_test_male$survived), positive = 'Yes')
conf_matrix_female <- confusionMatrix(factor(predicted_class_female), factor(data_test_female$survived), positive = 'Yes')

print("Male:")
print(conf_matrix_male)
cat("\n")
print("Female:")
print(conf_matrix_female)
```

#
We can see that indeed, at least within the testing groups, women did seem to survive at a higher proportion than men (24.8\% to 76.3\% in the testing set).  Print a summary of your trained model and interpret one of the fitted coefficients in light of the above disparity.  

```{r}
summary(model1)
```

*Student Input * 
The negative coefficient for sexmale suggests that being male significantly decreases the log odds of surviving compared to being female, holding all other variables constant. In practical terms, this means that, within the context of this model and data, men had lower odds of surviving the Titanic disaster than women. exp(-2.684206) ≈ 0.068. This odds ratio means that the odds of survival for men are approximately 6.8% of the odds of survival for women, given the same values for the other variables in the model (pclass, age, sibsp, and parch)

#

Now let's see if our model is *fair* across this explanatory variable.  Calculate five measures (as defined in class) in this question: the Overall accuracy rate ratio between females and males, the disparate impact between females and males, the statistical parity between females and males, and the predictive equality as well as equal opportunity between females and males (collectively these last two comprise equalized odds).  Set a reasonable $\epsilon$ each time and then comment on which (if any) of these five criteria are met.  


```{r}

accuracy_male <- conf_matrix_male$overall['Accuracy']
accuracy_female <- conf_matrix_female$overall['Accuracy']

tpr_male <- conf_matrix_male$byClass['Sensitivity']
tpr_female <- conf_matrix_female$byClass['Sensitivity']

fpr_male <- conf_matrix_male$byClass['Specificity']
fpr_female <- conf_matrix_female$byClass['Specificity']
fpr_male <- 1 - fpr_male
fpr_female <- 1 - fpr_female


ppv_male <- conf_matrix_male$byClass['Pos Pred Value']
ppv_female <- conf_matrix_female$byClass['Pos Pred Value']


oarr <- accuracy_female / accuracy_male
di <- ppv_female / ppv_male
spd <- (tpr_female + fpr_female) / 2 - (tpr_male + fpr_male) / 2
pe <- fpr_female - fpr_male
eo <- tpr_female - tpr_male

cat("Overall Accuracy Rate Ratio (OARR):", oarr, "\n")
cat("Disparate Impact (DI):", di, "\n")
cat("Statistical Parity Difference (SPD):", spd, "\n")
cat("Predictive Equality (PE):", pe, "\n")
cat("Equal Opportunity (EO):", eo, "\n")

```

*Student Input*.  

It is always important for us to interpret our results in light of the original data and the context of the analysis.  In this case, it is relevant that we are analyzing a historical event post-facto and any disparities across demographics identified are unlikely to be replicated.  So even though our model fails numerous of the statistical fairness criteria, I would argue we need not worry that our model could be misused to perpetuate discrimination in the future.  After all, this model is likely not being used to prescribe a preferred method of treatment in the future.  


#

Even so, provide a *philosophical* notion of justice or fairness that may have motivated the Titanic survivors to act as they did. Spell out what this philosophical notion or principle entails?

*Student Input* 
Justice as Need suggests that resources should be allocated based on individuals' specific needs. In the context of the Titanic, the decision to prioritize women and children can be seen as an application of this principle, recognizing their greater vulnerability in the disaster scenario. The notion implies that these groups had a more pressing need for the limited resource (lifeboats) to ensure their safety. 


